Depression and suicidality as costly signals of need

### 2021-04-29 Paper.rmd
More work on intro

<!-- ## Connection to Adversity -->

There is strong evidence that most cases of depression are caused by major negative life events [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. In longitudinal studies, depression onsets soon after a negative event [@KendlerEtAl1999] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after; @rosenstrom2017]. Additional support for the causal role of negative life events comes from the fact that the relationship between negative events and depression holds true when looking at events outside of one's control, showing that the connection is unlikely to be driven solely by individuals who are already depressed creating situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Additionally, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999].

@@

Other notable risk factors, like loss of a loved one or severe or prolonged illness, might seem less related to conflict. In these situations, however, the fitness costs that stem from reduced access to resources could be mitigated with help from others [@SugiyamaSugiyama2003] who might not be able, or want, to provide more investment than they already are. For this reason, many types of problems will eventually involve social conflict even if they did not start that way [@Hagen2003], something which is seen in the increasing evidence that loss of a loved one is often followed by increased family conflict [see @SymeHagen2021 for a brief review].

@@

With a cooperative species like our own, the ubiquity of conflicts of interests means that there will always be disagreement over the levels of investment in a cooperative endeavor and the division of the resulting benefits. In many such cases, individuals who were dissatisfied with the terms of cooperation could seek other partners [@hammerstein2016]. In many other cases, however, it would be difficult or impossible to switch partners. Spouses who were dissatisfied with their partner's investment in their new infant, for example, would not easily find a different partner to invest in that infant [@Hagen:1999]. Similarly, an adolescent who was dissatisfied with her parent's investment in her would not easily find other parents who were willing to invest more, nor could parents easily produce another adolescent. In ###these examples, and### many cooperative endeavors central to human biological fitness, each party has monopoly power over the benefits they bring to the endeavor -- no one is easily replaced [@Hagen2003].

@Hagen2003 proposed that physical aggression and depression were complementary strategies to resolve conflicts in interdependent relationships. @SellEtAl2009 found that physically formidable individuals were more prone to anger, prevailed more in conflicts of interest, and considered themselves entitled to better treatment. Physically weaker individuals, though, are not without options to resolve conflicts in their favor. An individual with monopoly power over the benefits she contributes to a critical cooperative endeavor can withhold those benefits, or put them at risk, until her partners change their behaviors in ways that benefit her. ### do we need a line about how depression results in delayed benefits here? ###

Withholding or risking one's contributions to a cooperative endeavor can be conceptualized as credibly signaling private information about one's need. Cooperation generates benefits over time. Thus, withholding or risking contributions incurs a *cost of delay* -- the more one values the endeavor, the greater the cost of delay. In a classic non-cooperative game theory model of bargaining over the division of benefits, @Rubenstein:1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information: in this case, each party knows exactly what division of benefits the other will accept, and can therefore make that offer immediately, avoiding the cost of delay. If valuations are private information, however, then immediate agreement might not be possible because each party has an incentive to deceptively report a low valuation in the hopes of getting more. A willingness to delay is a credible signal that one has a low valuation of the endeavor, and is therefore genuinely in need, because for those with a low valuation the benefit of a better offer outweighs the low cost of delay whereas for those for those with a high valuation the benefit of a better offer does not outweigh the higher cost of delay. Once valuations are known, the parties can reach an agreement on the division of benefits [@KennanWilson1993; @feinberg2005].

@Rosenthal1993 proposed a credible signaling model of what he termed "gambling-type" suicide attempts in which a probability of death less than 1 serves as the signal cost that allows signal receivers to accurately discriminate between those who are genuinely in need and those who are not.

Both the cost of delay of depression and the threat of death or serious injury that come from suicidal behavior allow for costly signaling strategies as they are relatively costly for those not facing adversity than those who are. In the case of depression, the large-scale reduction in one's normal routine of fitness enhancing behavior is relatively less costly for those whose situations result in a limited ability to obtain fitness-relevant benefits, with depression likely adding no additional costs in instances of extreme deprivation [@Hagen2003]. When it comes to suicide, the bulk of the costs associated with suicide attempts are only felt in times that result in death or severe bodily harm. For this reason, the way others view and respond to suicide attempts is likely to be strongly influenced by the probability that the attempt would result in death or severe injury [@Rosenthal1993]. Despite this difference, the previously mentioned logic still applies with increased risk of death being relatively more costly for those able to advance their fitness prospects in their day-to-day lives.

>>>>

<!-- ## Connection to Adversity -->
Almost all individuals will suffer adversity, such as injury or loss of material or social resources, at some point in their lives. Psychological pain, such as sadness and low mood, probably evolved to motivate victims of adversity to shift their attention to the causes of adversity so as to mitigate its negative fitness consequences and to learn to avoid future such adverse events [@thornhill1989; @andrews2010; @Nesse1990]. In many cases, social partners could help, and therefore signals of psychological pain, such as sad expressions and crying, probably evolved to indicate need [e.g., @balsters2013].

There is strong evidence that most cases of depression are caused by major negative life events [@Hammen2005; @Mazure1998; @devries2011; @ellsberg2008]. Compared to non-depressed individuals, those with depression report about twice as many instances of negative events than non-depressed individuals [@Mazure1998] and more negative events than those with schizophrenia and bipolar depression across multiple studies [@Paykel1994]. In longitudinal studies, depression onsets soon after a negative event [@KendlerEtAl1999] or coincides with periods where adversity is likely to increase prior to it [e.g., depression starting before a divorce rather than after; @rosenstrom2017]. Additional support for the causal role of negative life events comes from the fact that the relationship between negative events and depression holds true when looking at events outside of one's control, showing that the connection is unlikely to be driven solely by individuals who are already depressed creating situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. Additionally, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity, and that part of the heritability of depression stems from the heritability of negative events like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999]. For these and other reasons, most cases of depression are probably not due to mental illness but are instead functional instances of psychological pain [@Hagen:1999; @Hagen2003; @horwitz_loss_2007; @andrews2010; @SymeHagen2021].

@@

Other notable risk factors, like loss of a loved one or severe or prolonged illness, might seem less related to conflict. In these situations, however, the fitness costs that stem from reduced access to resources could be mitigated with help from others [@SugiyamaSugiyama2003] who might not be able, or want, to provide more investment than they already are. Indeed, there is increasing evidence that loss of a loved one is often followed by increased family conflict [see @SymeHagen2021 for a brief review]. Problems whose solutions require substantial investment or other changes on the part of social partners will therefore often involve social conflict even if they did not start that way [@Hagen2003]. When there are severe conflicts of interest, individuals have incentives to deceive, so "cheap" signals of need, such as sad expressions and crying, might not be believed.

@@

With a cooperative species like our own, the ubiquity of conflicts of interests means that there will always be disagreement over the levels of investment in a cooperative endeavor and the division of the resulting benefits. In many such cases, individuals who were dissatisfied with the terms of cooperation could seek other partners [@hammerstein2016]. In many other cases, however, it would be difficult or impossible to switch partners. Spouses who were dissatisfied with their partner's investment in their new infant, for example, would not easily find a different partner to invest in that infant [@Hagen:1999]. Similarly, an adolescent who was dissatisfied with her parent's investment in her would not easily find other parents who were willing to invest more, nor could parents easily produce another adolescent. In these examples, and many cooperative endeavors central to human biological fitness, each party has monopoly power over the benefits they bring to the endeavor -- no one is easily replaced [@Hagen2003].

@Hagen2003 proposed that physical aggression and depression were complementary strategies to resolve conflicts in interdependent relationships. @SellEtAl2009 found that physically formidable individuals were more prone to anger, prevailed more in conflicts of interest, and considered themselves entitled to better treatment. Physically or socially weaker individuals, though, are not without options to resolve conflicts in their favor. An individual with monopoly power over the benefits she contributes to a critical cooperative endeavor can withhold those benefits, or put them at risk, until her partners change their behaviors in ways that benefit her. Depression is characterized, in part by a profound loss of interest in virtually all activities [@dsm5], a symptom that is probably responsible for a sharp reduction in productivity [for review, see @SymeHagen2021]. Depression might therefore be a bargaining strategy for relatively powerless individuals in the wake of adversity and social conflict [@Hagen2003; @SymeHagen2021; see also @Watson:2002].

Withholding or risking one's contributions to a cooperative endeavor can be conceptualized as credibly signaling private information about one's need. Cooperation generates benefits over time. Thus, withholding or risking contributions incurs a *cost of delay* -- the more one values the endeavor, the greater the cost of delay. In a classic non-cooperative game theory model of bargaining over the division of benefits, @Rubenstein:1982 showed that two parties can come to an immediate agreement over division of benefits despite conflicts of interest if the parties' valuations of cooperation are not private information: in this case, each party knows exactly what division of benefits the other will accept, and can therefore make that offer immediately, avoiding the cost of delay. If valuations are private information, however, then immediate agreement might not be possible because each party has an incentive to request an amount greater than their actual valuation. A willingness to delay is a credible signal that one has a low valuation of the endeavor, and is therefore genuinely in need, because for those with a low valuation the benefit of waiting for a better offer outweighs the low cost of delay whereas for those for those with a high valuation the benefit of waiting for a better offer does not outweigh the higher cost of delay. Once valuations are known, the game reduces to the one in @Rubenstein:1982, and the parties can reach an agreement on the division of benefits [@KennanWilson1993; @feinberg2005].

@Rosenthal1993 proposed a credible signaling model of what he termed "gambling-type" suicide attempts. The vast majority of suicidal behavior does not result in death. In young adult women in the US, there are hundreds of attempts for every death (see Figure \@ref(fig:cdc)). Nevertheless, each attempt carries a non-zero risk of death. This cost allows social partners to accurately discriminate between those who are genuinely in need and those who are not [@Rosenthal1993]. In the case of depression, the large-scale reduction in one's normal routine of fitness enhancing behavior is relatively less costly for those whose situations result in a limited ability to obtain fitness-relevant benefits, with depression likely adding no additional costs in instances of extreme deprivation [@Hagen2003]. When it comes to suicide, the bulk of the costs associated with suicide attempts are only felt in times that result in death or severe bodily harm. For this reason, the way others view and respond to suicide attempts is likely to be strongly influenced by the probability that the attempt would result in death or severe injury [@Rosenthal1993]. Despite this difference, the previously mentioned logic still applies with increased risk of death being relatively more costly for those able to advance their fitness prospects in their day-to-day lives.
